---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
- Unzip and read the *activity.csv* file from working directory
- Convert date from character vector to a date vector
```{r}
library(dplyr)

data <- read.csv(unz("activity.zip", "activity.csv"), colClasses=c("numeric", "character", "numeric"))
## convert character to Date
data$date <- as.Date(data$date)

```



## What is mean total number of steps taken per day?

- Group steps by date variable
- Sum the step by 

```{r}
steps_sum <- group_by(data, date) %>% summarise( total_steps=sum(steps))
```

Plot histogram of steps taken per day: 

```{r}
hist(steps_sum$total_steps
     , breaks=4
     , xlab="Total number of steps taken per day")
```

Report mean and median: 
```{r}
mean(steps_sum$total_steps, na.rm=TRUE)
median(steps_sum$total_steps, na.rm=TRUE)

```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
