---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
- Unzip and read the `activity.csv` file from the zip file in the working directory
- Convert date variable from character vector to a date vector

```{r initialise, echo=TRUE}
## Load required libraries
library(dplyr)
library(lattice)
library(ggplot2)

hist_nb = 12

data <- read.csv(
    unz("activity.zip", "activity.csv"), 
    colClasses=c("numeric", "character", "numeric"))

## convert character to Date
data$date <- as.Date(data$date)

```


## What is mean total number of steps taken per day?

Summarise the data to calculate total steps taken per day

```{r}
steps_date <- group_by(data, date) %>% summarise( total_steps=sum(steps) )
```

Now plot histogram of steps taken per day: 

```{r}
hist(steps_date$total_steps
     , breaks=hist_nb
     , main = paste0("breaks=",hist_nb)
     , col = heat.colors(hist_nb, alpha = 1)
     , xlab="Total number of steps taken per day")
```

Calculate the mean and median value: 

```{r}
mean(steps_date$total_steps, na.rm=TRUE)
median(steps_date$total_steps, na.rm=TRUE)

```


## What is the average daily activity pattern?

Calculate the **mean** `steps` at 5-mins `interval` and plot the result

```{r}
steps_interval <- 
    group_by(data, interval) %>% summarise( mean=mean(steps, na.rm=TRUE))

ggplot(steps_interval, aes(interval, mean)) + 
    geom_line(color = "red", size = 0.5) + 
    labs(title = "Average Number of Steps Taken Across All Days"
         , x = "5-minute intervals (24h clock)"
         , y = "Average Number of Steps Taken")

```

The maximum *mean* steps
```{r}
steps_interval[which.max(steps_interval$mean), ]

```

**Observations**:

The time seriese graph indicate, the person's daily activity increases between 7:30 - 9:30 am and peaks around 8:35am. 


## Imputing missing values

Total number of NAs in the orginal dataset

```{r}
sum(is.na(data))
```

Create a new dataset, by replacing NA values from `steps` with mean (step) value at same `interval`.

```{r}
newData <- data

for( i in  1:nrow(newData)  ) {
    interval <- newData$interval[i]

    if( is.na(newData$steps[i]) ) {
        newData$steps[i] <- steps_interval[steps_interval$interval == interval, ]$mean
        
    }
    
}
```

Total number of NAs after imputing

```{r}
sum(is.na(newData))

```

Plot showing histogram after imputing

```{r}
steps_date <- 
    group_by(newData, date) %>% summarise( total_steps=sum(steps))

hist(steps_date$total_steps
     , breaks=hist_nb
     , main = paste0("breaks=",hist_nb)
     , col = heat.colors(hist_nb, alpha = 1)
     , xlab="Total number of steps taken per day")
```


The mean and median: 
```{r}
mean(steps_date$total_steps, na.rm=TRUE)
median(steps_date$total_steps, na.rm=TRUE)

```

**Observations**:

* After imputing the mean value remains the same. However, median values increseas slightly from orginal. Mean and Median are now equal. 
* On the new histogram's frequency density increase slightly where NA was replaced.



## Are there differences in activity patterns between weekdays and weekends?

Add a new weekday variable and partition the dataset into `weekday` and `weekend` using the `date` variable.

```{r}
## add two-level factoe variavle with weekday and weekend
newData$weekday <- 
    as.factor(ifelse(
        weekdays(newData$date) %in% c("Saturday","Sunday"),"Weekend", "Weekday")
        )

## Check how many weekday and weekends in the dataset
length(newData$weekday[newData$weekday =='Weekday'])
length(newData$weekday[newData$weekday =='Weekend'])

```

```{r}
# Calculate mean by weekday and interval
steps_date <- 
    group_by(newData, weekday, interval) %>% summarise( mean=mean(steps))

# xy panel plot
xyplot(steps_date$mean ~ steps_date$interval | factor(steps_date$weekday), 
       data=steps_date, 
       layout = c(1, 2), 
       type = 'l', 
       main="Average Number of Steps Taken\nAveraged Across All Weekdays or Weekend",
       xlab="5-Minute Interval (24h clock)",  
       ylab="Average Number of Steps Taken")

```

**Observation**: 
 The plot indicates that the person on avarage more active during the weekend. 
 